A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-19 text/introduction.tex
Revise -- getting close to the final form.

\subsection{Multi-stage Programming}

@@

Multi-stage programming enables us to generate and run fragments of code at runtime.  \AI{It sounds like the subject of ``generate and run'' is human being (``us'').  Something like ``Multi-stage programming offers linguistic support for dynamic code generation and execution.''?}
MetaOCaml~\cite{oleg2014} \AI{You need a space between a word and a citation.  Taha's original MetaOCaml should be cited, too.} provides features for multi-stage programming, which are brackets, escape, and run.  \AI{CSP?}
Brackets, written \verb|.< e >.| in MetaOCaml, makes a code value from \verb|e|.
Run, written \verb| run e | in MetaOCaml, runs a code value of \verb|e| and restore \verb|e|.
Escape, written \verb| .~ e | in MetaOCaml, expands a code value of \verb|e|.
Unlike run, escape is supposed to use in a fragment of code, that is, escape cannot be used as an alternative of run.
For example, the following MetaOCaml expression

\begin{verbatim}
        let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.
\end{verbatim}
evaluates to \verb|.< (fun x -> x + 1) 2 >.|. We can get the result with run.
\begin{verbatim}
        run (let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.)
\end{verbatim}
is reduced to 3.

% \subsubsection{CSPに関する補足説明}

Cross-stage persistence (CSP) is another primitive of multi-stage programming, which enable us to embed computed values into a code value.
This is an example of CSP.
\begin{verbatim}
        let plusone = fun x -> x + 1 in
        let y = plusone 2 in  .< y >.
\end{verbatim}
This program evaluates to \verb|.< 3 >.| not  \verb|.< plusone 2 >.|.
This is because the variable \verb|y| is introduced into a code fragment using CSP.
Therefore, the variable \verb|y| is embedded after it was calculated.
CSP is very important when we write practical programs 
because CSP enables us to use library functions in code fragments as in \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

The main application of multi-stage calculi is program optimization. 
A famous example is power functions.
Usual power functions take two arguments, which are the base and the exponent.
We can make a specialized power function for given exponents with multi-stage programming and 
optimize them by unrolling a loop in functions for given exponents.
The optimized power function for the exponent of \verb|3| looks like \verb|power3 = fun x -> x * x * x|.
\verb|power3| is faster than ordinary \verb|power| function because it contains no loop.
Multi-stage programming can optimize functions which take more than two arguments 
by generating a code fragment optimized to a given argument.

% \subsubsection{ベクトル計算も高速化できる}

Multi-stage programming can also optimize vector calculation.
For example, \verb|vadd| function, which takes two vectors and return the sum of them,
is implemented with a loop in many cases.
We can unroll \verb|vadd| function for a given vector length and optimize it with multi-stage programming.
\red{MetaOCamlで書いた長さ付きベクトルの例?}

% \subsubsection{生成した行列計算のコードは特定のサイズに特化しており、他のサイズでは使えない。
% このため、間違って組み合わせると容易にエラーを引き起こす。既存の型システムではこのエラーを防止できない}

However, there is a sever problem in functions which are optimized by multi-stage programming.
Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
Otherwise, we will get a Segmentation Fault error.
\AI{Why SEGV?  MetaOCaml will raise an exception.}
This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

% We call this problem \red{Specialized Code Problem}.

\AI{このイントロ、各段落はまあまあよいのですが、全体の流れが、ふたつ目の文から MetaOCaml という特定の多段階プログラミング言語の話をして、それから応用と問題点の話になりますが、ここが多段階プログラミング一般の話なのか MetaOCaml 固有の話なのかがぶれているように思います。まずは、多段階プログラミング一般の話を応用も含めてする。(他の多段階計算ができるシステムを無視しない。)それから、MetaOCaml の紹介と、その型システムの問題点の話をすべきでしょう。MetaOCaml の型システムでできることもふれておいた方がよいでしょう。}

@@

One natural idea to address this problem is the introduction of dependent types.
Dependent types are types which are dependent on values.
We can use dependent types for securer programming like Xi and Pfenning \cite{Xi98}.
For example, we can realize vectors with their sizes with them.
\begin{verbatim}
        Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor which takes the length of vectors, that is, 
\verb|Vector 3| is a type for vectors whose lengths are 3.
If \verb|vadd| function has the type of \verb|Vector n -> Vector n -> Vector n| for any positive number \verb|n|,
we can confirm the two arguments of \verb|vadd| function has the same length.
% \red{secure programming? finer grained typing? とか}

As we pointed out in the above section,
functions optimized with multi-stage programming can take only restricted values as arguments.
When you optimize \verb|vadd| function for the length of 5, you should it only for 5 length vectors.
We introduce dependent types into a multi-stage calculus
so that the type system can guarantee optimized functions are used properly.

In this paper, we design a new multi-stage calculus \LMD by 
merging dependent types into existing multi-stage calculus \LTP\cite{Hanada2014}.
\LMD is multi-stage calculus which contains dependent types.
\LMD makes multi-stage programming safer with the power of dependent types.
Our technical contribution are summarized as follows:
\begin{itemize}
        \item we give the formal definition of \LMD with its syntax, 
        type system which contains type equivalence, full, reduction, and staged reduction;
        \item for the full reduction, we prove subject reduction, strong normalization, and confluence
        \item fot the staged reduction, we prove unique decomposition and progress.
\end{itemize}

@@

The organization of this paper is the following.
Section 2 gives an informal overview of \LMD.
Section 3 defines \LMD and
Section 4 shows the properties of \LMD including Unique Decomposition \AI{There are more important properties that should be mentioned!  In fact, the properties should be mentioned earlier.}.
Section 5 discusses future works.
Finally, Section 6 discusses related works.
\AI{Use $\backslash$label and $\backslash$ref!}
\AI{``Related work'' and ``future work'' (they are uncountable).}
\AI{Related work is discussed in Section 5?  Section 6 is the section for conclusion, which includes future work.}

>>>>

\subsection{Multi-stage Programming and MetaOCaml}

@@

\AI{前の版は，まだ MetaOCaml specific な話と，MSP 一般の話が交互に出てきて
わかりづらかったと思います．}

Multi-stage programming makes it easier for programmers to implement
generation and execution of code at run time by providing language
constructs for composing and running pieces of code as first-class
values.  A promising application of multi-stage programming is
(run-time) code specialization, which is to generate program code
specialized to partial inputs to the program and such applications are
studied in the literature~\cite{taha2007gentle}  \AI{We should list
applications of multi-stage programming, not necessarily of MetaOCaml.}

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with constructs for multi-stage
programming, including brackets, escape, run, and cross-stage
persistence (CSP) .  Brackets, written \verb|.< e >.| in MetaOCaml,
makes a value that represents the code (or abstract syntax tree) of
expression \verb|e|; and escape, written \verb|.~ e| and is supposed
to appear inside brackets, evaluates \verb|e| to a code value and
expands the code into the surrounding code.
For example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.
\end{verbatim}
evaluates to \verb|.< (fun x -> x + 1) 2 >.|.  
Run, written \verb|run e|, evaluates \verb|e| to a code value
and executes it.  For example,
\begin{verbatim}
   run (let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.)
\end{verbatim}
yields \verb|3| (here, escape connects tighter than application).
Finally, CSP is a primitive to embed values (not necessarily code
values) into a code value.  In MetaOCaml, CSP is implicitly applied to
occurrences of variables defined outside brackets.  For example,
\begin{verbatim}
   let plusone = fun x -> x + 1 in
   let y = plusone 2 in  .< y * 2 >.
\end{verbatim}
evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
This is because the variable \verb|y| is bound to an integer \verb|3|
before \verb|.<y>.| is evaluated and the value of \verb|y| is embedded
into the code value.  In MetaOCaml, CSP can be applied to (variables of) any type.  CSP is practically important because it enables us to use library functions in brackets, e.g., \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

% As we mentioned earlier, a main application of multi-stage
% calculi is program optimization. 
% A famous example is power functions.
% Usual power functions take two arguments, which are the base and the exponent.
% We can make a specialized power function for given exponents with multi-stage programming and 
% optimize them by unrolling a loop in functions for given exponents.
% The optimized power function for the exponent of \verb|3| looks like \verb|power3 = fun x -> x * x * x|.
% \verb|power3| is faster than ordinary \verb|power| function because it contains no loop.
% Multi-stage programming can optimize functions which take more than two arguments 
% by generating a code fragment optimized to a given argument.

% 多段階計算の型理論の既存研究の紹介

MetaOCaml is also equipped with a type system for safe code generation
and execution.  The notion of code types is introduced to prevent code
values that represent ill-typed expressions from being generated.  For
example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone "foo" >.
\end{verbatim}
is rejected by the MetaOCaml typechecker as \verb|plusone| is given
type \verb|(int -> int)| \verb|code| but it is used in the code context that
requires a function from \verb|string| is expected.  Ensuring safety
for \verb|run| is more challenging because basic code types do not
guarantee code values to be closed.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem and developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values with a type soundness theorem.

However, the type system, which is based on Hindley--Milner
polymorphism~\cite{}, is not strong enough to guarantee invariants
beyond simple types.  For example, Kiselyov~\cite{} demonstrates
specialization of vector/matrix computation with respect to the sizes
of vectors and matrices but such specialized functions can be applied
to vectors and matrices of different sizes.

% Multi-stage programming can also optimize vector calculation.
% For example, \verb|vadd| function, which takes two vectors and return the sum of them,
% is implemented with a loop in many cases.
% We can unroll \verb|vadd| function for a given vector length and optimize it with multi-stage programming.

% % 多段階計算で生成したコードは特殊化されているが故に問題も多い

% However, there is a sever problem in functions which are optimized by multi-stage programming.
% Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
% For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
% Otherwise, we will get a exception.
% This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specilization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector|$n$\verb| -> Vector |$n$\verb|) code|.

% As we pointed out in the above section,
% functions optimized with multi-stage programming can take only restricted values as arguments.
% When you optimize \verb|vadd| function for the length of 5, you should it only for 5 length vectors.
% We introduce dependent types into a multi-stage calculus
% so that the type system can guarantee optimized functions are used properly.
% Although there is a trial to combine multi-stage programming and dependent type by Brady and Hammond~\cite{brady2006dependently},
% they didn't give a formal definition and properties for the calculus.

In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014}---which
formalizes the four constructs of MetaOCaml in a way slightly
different from MetaOCaml---with dependent types and study its
properties.  We base our work on \LTP because its type system and
semantics are arguably simpler than
\(\lambda^\alpha\)~\cite{taha2003environment}, which is more faithful
to the design of MetaOCaml.  Our technical contribution are summarized
as follows:
\begin{itemize}
\item We give the formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.  Section
\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.
